---
title: "Introduction to survival analysis with fixedtimeanalyzR"
author: "Lukas D. Sauer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: >
  This vignette is a basic introduction to the mathematical idea behind and the
  functionality of the fixedtimeanalyzR package. This package offers alternative
  methods for comparing censored survival data in cases where a log-rank test is
  not applicable.
vignette: >
  %\VignetteIndexEntry{Introduction to survival analysis with fixedtimeanalyzR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When comparing censored survival data of two groups, it is most common to use a log-rank test. However, it is not always possible to apply a log-rank test: Imagine that you would like to compare the survival rate of your lung cancer patients to the survival rate from a closed-access publication. In this case, alternative methods are necessary for comparing survival rates. John P. Klein et al. developed test statistics for comparing the numerical values of survival rates, which are implemented in this package.

# Example 1: Comparing a dataset to a numerical value

Consider the **NCCTG Lung Cancer Data** from the `survival`package.
```{r}
lung
data(cancer, package="survival")
```



Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
