---
title: "Introduction to survival analysis with fixedtimeanalyzR"
author: "Lukas D. Sauer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: >
  This vignette is a basic introduction to the usage of the fixedtimeanalyzR
   package. This package offers alternative methods for comparing censored 
   survival data in cases where a log-rank test is not applicable.
vignette: >
  %\VignetteIndexEntry{Introduction to survival analysis with fixedtimeanalyzR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When comparing censored survival data of two groups, it is most common to use a log-rank test. However, it is not always possible to apply a log-rank test: Imagine that you would like to compare the survival rate of your lung cancer patients to the survival rate from a closed-access publication. Whenever the full survival data set is not available, a log-rank test is not applicable.

In this case, alternative methods are necessary for comparing survival rates. John P. Klein et al.^[Klein, J.P., Logan, B., Harhoff, M. and Andersen, P.K. (2007), [Analyzing survival curves at a fixed point in time](https://doi.org/10.1002/sim.2864). Statist. Med., 26: 4505-4519.] developed test statistics for comparing the numerical values of survival rates, which are implemented in this package.

# Examples
Setup instructions can be found in the [installation vignette](installation.html). In the examples, we will use the following packages:

```{r, message=FALSE}
library(tidyverse)
library(survival)
library(fixedtimeanalyzR)
```

## Example 1: Comparing one group to a numerical value

Consider the *NCCTG Lung Cancer Data* from the `survival` package.
```{r, results=FALSE}
data(cancer, package="survival")
lung
```
```{r, echo=FALSE, results='asis'}
knitr::kable(head(lung, 5))
```
It contains the two variables necessary for **fixedtimeanalyzR** to work:

- `time`, specifying the time to event or censoring for each row,
- `status`, specifying whether the event in the row was an actual event or merely the censoring of an individual. It may have the same codes as the `event` variable of the `Surv` object from the `survival` package. This means that we can have the following codes: 0=censored and 1=event, or 1=censored and 2=event or FALSE=censored and TRUE=event.


Apart from these two variables, a data set may contain a variable `group`, specifying the treatment group that each individual belongs to. This will be discussed in the next example.

Let's suppose that we want to compare the survival rate to a given value. We claim that the survival rate of the `lung` cancer patients after 365 days is different from 60% (with a standard error of 5%).

**fixedtimeanalyzR** provides different tests for this comparison. Each test is a chi-squared test with one degree of freedom, but they differ in the choice of their test statistics. For this example, we choose a clog-transformed statistic, the other statistics will be explained further down.

Comparison is now as easy as writing
```{r}
clog.test(lung, surv_KM=0.6, se_KM=0.05, t=365)
```
where `clog.test` is the statistical test based on the clog-transformed test statistic, `surv_KM` is the Kaplan-Meier survival rate, `se_KM` is the corresponding standard error, and `t` is the time at which we want to compare.


We obtain an estimated survival rate of approximately 41% for the `lung` data set, and p-value of 0.003, meaning that the difference between our data set and the reference value of 60% is statistically significant with respect to a significance level of 5%.

The function output is a list of class `htest`, containing a lot of information on the test. For example, use `clog.test(lung, surv_KM=0.6, se_KM=0.05, t=365)$p.value` to get the p-value and `clog.test(lung, surv_KM=0.6, se_KM=0.05, t=365)$statistic` in order to get the value of the test statistic.

## Example 2: Comparing two groups to each other

Instead of comparing two groups via a log-rank test, you can also use this package to compare two groups at a fixed point in time, using Klein et al.'s methods. Consider the  data set `aml`, containing time to censoring/time to death for patients with acute myelogenous leukemia.
```{r, results=FALSE}
data(cancer, package="survival")
aml
```
```{r, echo=FALSE, results='asis'}
knitr::kable(head(aml, 5))
```

According to the [`survival` package documentation](https://cran.r-project.org/web/packages/survival/survival.pdf), the variable `x` describes "whether the standard course of chemotherapy [was] extended (’maintainance’) for additional cycles".

In order to compare these two treatment groups, we need to rename the variable `x`.
```{r}
aml %>% rename(group = x) -> amll
```
Now, we're all set. We can compare the survival rate at `time==20` for the two treatment groups using the same test as in example 1.
```{r}
clog.test(amll, t=20)
```
The survival rate estimate at  `time==20` of the "maintained" group is 72%, which is higher than the estimate for the group for which chemotherapy was not maintained (58%). However, as consequence of the large standard errors, this difference is not statistically significant with respect to a significance level of 5%.

## Example 3: Comparing two numerical values

Finally, it is also possible to compare the numerical values of two survival rates. Let us suppose that we want to compare the survival rate 50% (SE 5%) to the rate 60% (SE 6%). This can be achieved by typing
```{r}
clog.test(surv_KM=c(0.5, 0.6), se_KM=c(0.05,0.06))
```
Again, we see that the difference of these two rates is not statistically significant with respect to a significance level of 0.05.

## Example 4: Passing a survfit object

Internally, **fixedtimeanalyzR** coerces a the data frame `data` into an object of class `survfit`. Instead of passing a data frame, it is also possible to directly pass a `survfit` object to the tests. This may have two advantages:

* You gain code speed when running a lot of tests.
* You have the possibility of choosing your own variable names instead of `time`, `status` and `group`.

Have a look at the following data frame from the survival package, containing the data of 2982 primary breast cancer patients from the *Rotterdam tumor bank*.

```{r, results=FALSE}
data(cancer, package="survival")
rotterdam
```

```{r, echo=FALSE, results='asis'}
knitr::kable(head(rotterdam, 5))
```

We will set up a `survfit` object with time to death/time to censoring `dtime`, status variable `death` where 0=alive and 1=dead and the application of chemotherapy `chemo` as a grouping variable.
```{r}
rotterdamfit = survfit(formula = Surv(dtime, death) ~ chemo, data=rotterdam)
```
Now, this object can be passed to our test as before. We will test for the survival rate after three years, i.e. 1095 days.
```{r}
clog.test(rotterdamfit, t=1095)
```


# Background: Different test statistics

Apart from `clog.test`, four more tests relying on different test statistics are implemented in **fixedtimeanalyzR**. Each test statistic is taken from Klein et al.'s article.

The test `naive.test` simply relies on the test statistic

$$ X_1^2=\frac{(S_1(t)-S_2(t))^2}{\mathrm{se}(t)_1^2 + \mathrm{se}(t)_2^2},$$
where $S_i(t)$ are the survival rates at time $t$ for each group $i$, and $se_i(t)$ are the respective standard errors.

The test `logtra.test` relies on the log-transformed test statistic

$$ X_2^2=\frac{(\log(S_1(t))-\log(S_2(t)))^2}{\sigma_1(t)^2 + \sigma_2(t)^2}$$
with the quotient $\sigma_i(t)=\frac{\mathrm{se}_i(t)}{S_i(t)}$. The square of this quotient can be calculated by Greenwood's formula, we maintain this notation for the sake of better readability.

The test `clog.test` relies on the the clog-transformed test statistic

$$ X_3^2 = \frac{(\log(-\log(S_1(t)))-\log(-\log(S_2(t))))^2}{( \sigma_1(t)/\log(S_1(t)) )^2+( \sigma_2(t)/\log(S_2(t)) )^2}.$$
The test `asinsqrt.test` relies on the following test statistic transformed by arc sine and square root:
$$ X_4^2 = \frac{\left( \arcsin(\sqrt{S_1(t)})-\arcsin(\sqrt{S_2(t)})\right)^2}{\nu_1(t)+\nu_2(t)}$$
where $\nu_i(t) = \frac{S_i(t)\sigma_i(t)^2}{ 4(1-S_i(t))}$.

Finally, the test `logit.test` relies on the logit-transformed test statistic
$$ X_5^2 = \frac{\left( \log\frac{S_1(t)}{1-S_1(t)} - \log\frac{S_2(t)}{1-S_2(t)} \right)^2}{\frac{\sigma_1(t)^2}{1-S_1(t)^2} + \frac{\sigma_2(t)^2}{1-S_2(t)^2}}.$$
For example, we could compare the p-values of the different tests:
```{r}
data = lung
surv_KM = 0.6
se_KM = 0.05
t=365
p.vals = c(naive.test(data, surv_KM, se_KM, t)$p.value,
           logtra.test(data, surv_KM, se_KM, t)$p.value,
           clog.test(data, surv_KM, se_KM, t)$p.value,
           asinsqrt.test(data, surv_KM, se_KM, t)$p.value,
           logit.test(data, surv_KM, se_KM, t)$p.value)
names(p.vals) = c("Naive test",
                  "log test",
                  "clog test",
                  "asin&sqrt test",
                  "logit test")
p.vals
```

